Global patterns and drivers of vertebrate verticality

### 2025-02-25 verticality_future_mkd.Rmd
figures and methods updates

To examine spatial patterns of assemblage verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality: $SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool. For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell by conducting 100 simulations with random samples matching the observed richness. Highly positive SES verticality scores represent assemblages having much higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.

@@

We used 10 variables to represent variation in environmental conditions. Climate data were represented with six predictors: (i) mean daily maximum air temperature of the warmest month, (ii) mean daily minimum air temperature of the coldest month, (iii) temperature seasonality, (iv) precipitation of the wettest month, (v) precipitation of the driest month, and (vi)  precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. All climatic variables were extracted from CHELSA v2.1 (30 arc-second resolution) for both current (1981-2010) and future (2071-2100) periods, the later represented by an ensemble of the five general circulation models in CHELSA CMIP6 ISIMIP3 data projected under the Shared Socioeconomic Pathway (SSP) 585. Vegetation data included (vii) canopy height [@langHighresolutionCanopyHeight2023], (viii) vegetation density [@crowtherMappingTreeDensity2015], and (ix) vegetation complexity (the product of canopy height and vegetation density). Additionally, we included (x) historical climate velocity (0.25° resolution), which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and influences global patterns of verticality [@oliveiraVerticalStratificationInfluences2019]. All environmental data were resampled to the spatial resolution and projection of the species data.

@@

```{r fig.id="richness-mods", fig.width = 160, fig.height=115.6, fig.cap.pre="Fig. ", fig.cap="Slopes for linear models of richness as a function of SES or mean verticality across all biomes and for individual biomes. Hollow circles indicate non-significant relationships (p ≥ 0.05) and filled circles indicate significant relationships. R^2^ values are represented by the size of the circle. Purple = all biomes, gray = boreal biomes, orange = temperate biomes, green = tropical biomes. Flooded GS = Flooded grassland and savannas, Mediterranean FWS = Mediterranean forests, woodlands, and scrub, Montane GS = Montane grasslands and shrublands, Temperate BMF = Temperate broadleaf and mixed forests, Temperate CF = Temperate conifer forests, Temperate GSS = Temperate grasslands, savannas, and shrublands, Tropical CF = Tropical and subtropical conifer forests, Tropical DBF = Tropical and subtropical dry broadleaf forests, Tropical GSS = Tropical and subtropical grasslands, savannas, and shrublands, Tropical MBF = Tropical and subtropical moist broadleaf forests"}

include_graphics("figures/main_figs/fig5_vert_rich_lm_slopes.png", error = F)
```

>>>>

To examine spatial patterns of assemblage verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality: $SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool. For each vertebrate class, we defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell based on 100 simulations. For each simulation, we randomized verticality within the grid cell by randomly drawing the same number of verticality scores from the respective species pool as the observed richness in the cell. Highly positive SES verticality scores represent assemblages having much higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.

@@

We used 11 variables to represent variation in environmental conditions. Climate data were represented with seven predictors: (i) mean daily maximum air temperature of the warmest month (Tmax warm), (ii) mean daily minimum air temperature of the coldest month (Tmin cold), (iii) temperature seasonality (Temp sea), (iv) precipitation of the wettest month (Precip wet), (v) precipitation of the driest month (Precip dry), (vi)  precipitation seasonality, and (vii) mean monthly precipitation of the warmest quarter [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. All climatic variables were extracted from CHELSA v2.1 (30 arc-second resolution) for both current (1981-2010) and future (2071-2100) periods, the later represented by an ensemble of the five general circulation models in CHELSA CMIP6 ISIMIP3 data projected under the Shared Socioeconomic Pathway (SSP) 585. Vegetation data included (viii) canopy height [@langHighresolutionCanopyHeight2023], (ix) vegetation density [@crowtherMappingTreeDensity2015], and (x) vegetation complexity (the product of canopy height and vegetation density). Additionally, we included (xi) historical climate velocity (0.25° resolution), which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and influences global patterns of verticality [@oliveiraVerticalStratificationInfluences2019]. All environmental data were resampled to the spatial resolution and projection of the species data.

@@

```{r fig.id="cor-plot", fig.width = 180, fig.height=91.35, fig.cap.pre="Fig. ", fig.cap=" Pearson correlation coefficients per biome for relationships between richness and SES verticality (x-axis) and SES verticality and canopy height (y-axis) for a) birds, b) mammals, c) reptiles, and d) amphibians. Colors indicated biome. Black rings indicate significance of the correlation. The outer ring indicates significance of the correlation between verticality and canopy height and the inner ring indicates significance of the correlation between richness and verticality."}

include_graphics("figures/main_figs/fig4/fig4_correlations_annotated.png", error = F)
```
